Seismic parameter estimation and the Canadian crust

### 2013-11-18 paper/draftBP.tex
half intro complete

%% Seismic studies of the Canadian continental crust predominately analyze specific geological features or regions rather than taking a comprehensive and comparative inter-regional analysis. The primary reason for this is the poor resolution afforded by the seismic networks currently and historically deployed across Canada. The Canadian continental landmass is composed of at least fifteen large geological provinces as recognized by the Geological Survey of Canada (GSC). Each of these regions is itself complex and heterogeneous and often larger than most European nations. On top of this there is poor seismic coverage, roughly one seismic station per 25,000 $km^2$. Many of these stations are clustered near areas of geologic interest such as the Cascadia Subduction zone or population and research centres such as Southern Ontario or areas of significant resource interest like the diamondiferous Great Slave Lake area. This leaves vast areas of the Canadian landmass completely unsampled. Despite these limitations, a low resolution comprehensive and comparative study is still feasible for some of the geological regions comprising the Canadian continental crust. Direct comparison between the bulk average seismic properties of geological regions and between aggregated regions and global averages provide some new insight into the crustal composition of the Canadian landmass. The accumulated dataset also affords investigations into variations of bulk crustal parameters such as Poisson's Ratio and crustal thickness with age and tectonic environment as well as some basic statistical data on interesting crustal features.

@@

Meissner, R., (Ed.) (1986) \it{The Continental Crust: A Geophysical Approach}, 426, Academic, San Diego, Calif.

>>>>

The Lithoprobe project has provided researches with a rich dataset from which detailed cross-sections have been derived. These cross sections form a near continuous transect from the Pacific to the Atlantic oceans have and enabled many insights into crustal structure and genesis. Recent examples include Cook et al [2009] analysis of the Moho where they present the Moho as having varied and complex origins and as evolving through thermal and pressure changes, deformation and fluid migration. In another recent Lithoprobe study, Hammer and Clowes [2011] discuss broad structural characteristics of the Canadian crust. They note a remarkable consistency in orogenic style such as the similarity between geolgical structures contained within Archean cratons and their boundaries as well as those observed in Proterozoic and Phanerozoic orogens. Also noted is the striking consistency in moho depth, which is imaged as lying between 33km and 43km depth along the full 20,000km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. and Hammer and Clowes raise the possibility that in some situations the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary. A final example of a Canada wide crustal study utilizing Lithoprobe data is the Lith5.0 revised crustal model presented by Perry, Eaton and Forte [2002]. This work updates the Crust5.1 model [Mooney et al, 1998] to include the new data from Lithoprobe and introduces broad crustal features of the Canadian crust such as the relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick (>42.5km) Phanerozoic crust in the south and west of the continental craton. These studies and other Canada wide studies employ active source Lithoprobe data which records higher frequency (10-40hz) {\it P}-wave velocity whereas this study uses lower frequency (0.4-3hz) passive source scattering techniques which provides estimates of the $V_P/V_S$ ratio and images relatively larger scale structure. Studies of the Canadian crust employing a similar teleseismic receiver function method as this work include a study of the Great Slave Lake shear zone [Eaton and Hope, 2002] which utilize receiver functions, shear-wave splitting results and gravity data to characterize the shear zone. Finally, three teleseismic studies discussed in greater detail in later sections are presented in chronological order: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen [Eaton et al., 2005], a structural and compositional characterization of the Superior Province [Darbyshire et al. 2006] and a crustal evolution analysis of the Hudson Bay region [Thompson et al., 2010]. This work differs from these studies in that it focuses on interregional comparisons and Canada wide structural characteristics, as does some of the Lithoprobe literature, but also on Canadian bulk crustal composition.

While there is a great diversity and body of seismic research on the Canadian crust those employing similar teleseismic methods and those encompassing the whole Canadian landmass require mention here. Recent teleseismic receiver function studies in the Candian Shield include

In the following sections we describe the data and methods that we employ to interrogate the Canadian crust. We then proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three published for matching seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the main dataset in two sections, one focusing on crustal discontinuities and structure and the other on regional variation and characterization. Finally we conclude with a discussion of implications of the results for our understanding of the Conrad discontinuity and bulk crustal composition of the Canadian Crust.
%% Seismic studies of the Canadian continental crust predominately analyze specific geological features or regions rather than taking a comprehensive and comparative inter-regional analysis. The primary reason for this is the poor resolution afforded by the seismic networks currently and historically deployed across Canada. The Canadian continental landmass is composed of at least fifteen large geological provinces as recognized by the Geological Survey of Canada (GSC). Each of these regions is itself complex and heterogeneous and often larger than most European nations. On top of this there is poor seismic coverage, roughly one seismic station per 25,000 $km^2$. Many of these stations are clustered near areas of geologic interest such as the Cascadia Subduction zone or population and research centres such as Southern Ontario or areas of significant resource interest like the diamondiferous Great Slave Lake area. This leaves vast areas of the Canadian landmass completely unsampled. Despite these limitations, a low resolution comprehensive and comparative study is still feasible for some of the geological regions comprising the Canadian continental crust. Direct comparison between the bulk average seismic properties of geological regions and between aggregated regions and global averages provide some new insight into the crustal composition of the Canadian landmass. The accumulated dataset also affords investigations into variations of bulk crustal parameters such as Poisson's Ratio and crustal thickness with age and tectonic environment as well as some basic statistical data on interesting crustal features.

@@

Meissner, R., (Ed.) (1986) {\it The Continental Crust: A Geophysical Approach}, 426, Academic, San Diego, Calif.
